---
title: "Week 1 Assignment: Code Base"
author: "Dillon Leeper"
format:
  html:
    toc: true
    code-fold: show
editor_options:
  markdown:
    wrap: sentence
---

## Introduction

This assignment uses the [Online Retail dataset from the UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/352/online+retail), which contains transactions from a UK-based e-commerce store between December 2010 and December 2011.
The goal is to load the original data reproducibly, select and rename useful columns, and create fields that make cancellations and transaction values easier to analyze.

## Approach

For this assignment, I plan to use the Online Retail dataset from the UCI Machine Learning Repository.
The dataset contains transaction data from a UK-based online retailer and includes information such as invoice number, product code and description, quantity, invoice date, unit price, customer ID, and country.

I chose this dataset because I work with e-commerce data regularly, so I am already familiar with the type of information it contains.
I think it will be useful to work with familiar business data while learning how to load, clean, and transform data in R.

For the assignment, I plan to select a smaller set of useful columns, rename some of them to make them easier to understand, and create a field that identifies whether an order was cancelled based on the invoice number.

One data challenge I expect is handling cancelled orders.
Any InvoiceNo that starts with C represents a cancellation.
If those transactions are not identified and handled correctly before analysis, they could distort sales-related results.

**Dataset source:** [UCI Machine Learning Repository, Online Retail](https://archive.ics.uci.edu/dataset/352/online+retail)

**Dataset:** Online Retail dataset

## Load the data

The original dataset is stored in a ZIP archive on UCI's website.
The code below downloads that archive to a temporary directory, extracts the Excel workbook, and loads it into R.
This approach avoids relying on a file stored only on my computer.

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
options(width = 100)
```

```{r load-data}
data_url <- "https://archive.ics.uci.edu/static/public/352/online+retail.zip"

zip_file <- tempfile(fileext = ".zip")
data_directory <- tempfile(pattern = "online-retail-")
dir.create(data_directory)

download.file(data_url, zip_file, mode = "wb", quiet = TRUE)
unzip(zip_file, exdir = data_directory)
excel_file <- list.files(
  data_directory,
  pattern = "[.]xlsx$",
  full.names = TRUE
)[1]
online_retail_raw <- read_excel(excel_file)
```

The source data contains `r format(nrow(online_retail_raw), big.mark = ",")` rows and `r ncol(online_retail_raw)` columns.

```{r inspect-source}
names(online_retail_raw)
head(online_retail_raw)
```

## Transform the data

The transformed data frame keeps seven useful source columns and gives them consistent, descriptive names.
It also adds three fields:

- `is_cancelled` identifies invoices whose number begins with `C`.
- `line_value_gbp` calculates quantity multiplied by unit price.
- `transaction_type` provides a readable label for completed purchases and cancellations.

```{r transform-data}
online_retail <- data.frame(
  invoice_number = as.character(online_retail_raw$InvoiceNo),
  product_description = as.character(online_retail_raw$Description),
  quantity = as.integer(online_retail_raw$Quantity),
  invoice_date = as.POSIXct(online_retail_raw$InvoiceDate),
  unit_price_gbp = as.numeric(online_retail_raw$UnitPrice),
  customer_id = as.character(online_retail_raw$CustomerID),
  country = as.character(online_retail_raw$Country),
  stringsAsFactors = FALSE
)
online_retail$is_cancelled <- startsWith(
  toupper(online_retail$invoice_number),
  "C"
)
online_retail$line_value_gbp <-
  online_retail$quantity * online_retail$unit_price_gbp
online_retail$transaction_type <- ifelse(
  online_retail$is_cancelled,
  "Cancellation",
  "Completed purchase"
)
```

## Verify the result

The following output confirms the structure of the transformed data frame and summarizes its transaction types.
Keeping cancellations clearly labeled is important because their negative quantities and values could otherwise distort sales totals.

```{r verify-result}
str(online_retail)
head(online_retail)
transaction_summary <- data.frame(
  transaction_type = names(table(online_retail$transaction_type)),
  row_count = as.integer(table(online_retail$transaction_type)),
  row.names = NULL
)
transaction_summary$percent_of_rows <- round(
  100 * transaction_summary$row_count / sum(transaction_summary$row_count),
  2
)
transaction_summary
```

```{r validation-checks}
stopifnot(
  nrow(online_retail) == nrow(online_retail_raw),
  ncol(online_retail) == 10,
  all(online_retail$transaction_type %in%
        c("Cancellation", "Completed purchase")),
  all(online_retail$is_cancelled ==
        startsWith(toupper(online_retail$invoice_number), "C"))
)
```

## Conclusions

The final data frame is a smaller and more understandable version of the original dataset.
The analysis identified 9,288 cancellation records, representing 1.71% of all transaction rows.
It retains the transaction, product, customer, date, price, and country information needed for later analysis while explicitly identifying cancelled transactions and calculating each row's value in British pounds.

As a next step, I could verify how missing customer and product descriptions should be handled, separate valid product sales from returns or data-entry corrections, and aggregate completed revenue by month or country.
I would keep cancellations separate from completed purchases so that summaries do not accidentally treat both transaction types as ordinary sales.

## Data source

Chen, D.
(2015).
*Online Retail* \[Dataset\].
UCI Machine Learning Repository.
<https://doi.org/10.24432/C5BW33>

## AI Assistance

AI assistance: ChatGPT was used during development of this assignment.
The complete interaction transcript is included in `ai-transcript.md`.

**AI tool citation:** OpenAI.
(2026).
ChatGPT (GPT-5.6 Sol) \[Large language model\].
Accessed August 30, 2026.
