---
title: "Assignment 3A: Global Baseline Estimate — Approach & Code Base"
subtitle: "DATA 607 · Week 3"
author: "Dillon Leeper"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
    css: styles.css

execute:
  warning: false
  message: false

editor: visual
---

## Overview

Most recommender systems rely on personalized algorithms such as content-based filtering or item-item collaborative filtering. The **Global Baseline Estimate** is a useful, simpler alternative: a non-personalized recommender that predicts a rating using only the overall rating distribution, how a given item tends to be rated relative to everyone else, and how a given user tends to rate relative to everyone else. It serves as a benchmark that any personalized algorithm should be expected to beat.

This document implements a Global Baseline Estimate recommender in R using the movie ratings survey data collected for Assignment 2A — the same 7-respondent, 6-movie dataset stored in PostgreSQL and documented in `week02_2a_approach.qmd`.

## The Algorithm

For a user $u$ and item (movie) $i$, the baseline prediction is:

$$
\text{baseline}(u, i) = \mu + b_u + b_i
$$

where:

- $\mu$ — the global mean rating across all users and all movies
- $b_i$ — the item bias: how much movie $i$'s average rating deviates from $\mu$
- $b_u$ — the user bias: how much user $u$'s ratings deviate from $\mu$, after accounting for the items they rated

The result is typically clipped to the valid rating scale, since $\mu + b_u + b_i$ can fall outside the original bounds.

## Planned Approach

### 1. Data Preparation

Import the survey data and reshape it from wide format (one row per respondent, one column per movie) into long/tidy format with `tidyr::pivot_longer()`, so each row represents a single (user, movie, rating) observation. Missing/unrated cells will be filtered out rather than treated as zero ratings.

### 2. Compute the Global Mean

Calculate $\mu$ as the mean of every recorded rating in the long-format data frame, ignoring missing values.

### 3. Compute Item (Movie) Bias

For each movie, compute:

$$
b_i = \overline{\text{rating}_i} - \mu
$$

grouping by movie and averaging its ratings, then subtracting the global mean.

### 4. Compute User Bias

For each user, correct for item bias before averaging: subtract both $\mu$ and the relevant movie's $b_i$ from each of that user's ratings, then average the residuals:

$$
b_u = \text{mean}(\text{rating}_{u,i} - \mu - b_i)
$$

### 5. Compute Baseline Predictions

Combine $\mu$, $b_u$, and $b_i$ for every (user, movie) pair — including pairs the user has not yet rated — to produce a predicted rating, clipped to the valid scale (1–5).

### 6. Generate Recommendations

For each user, rank the movies they have not yet rated by predicted baseline score and surface the top-ranked titles as recommendations.

## R Implementation

```{r}
#| label: setup

library(tidyverse)
```

### Dataset

This reproduces the exact 7-respondent × 6-movie survey table from Assignment 2A (`week02_2a_approach.qmd`), where `NA` marks "haven't seen it" / unanswered cells — never a 0 rating.

```{r}
#| label: build-dataset

ratings_wide <- tribble(
  ~user,          ~`Everything Everywhere All at Once`, ~`The Batman`, ~Hereditary, ~Obsession, ~`The Odyssey`, ~`Mean Girls`,
  "Eduardo",       4,  3, NA, NA,  5, NA,
  "Annie",         4,  3, NA, NA, NA,  5,
  "James",         2,  2, NA, NA,  4,  3,
  "JOYEEEEEEEE",  NA, NA, NA,  5,  4, NA,
  "Chris",         5, NA,  3,  5, NA,  5,
  "Garrett",      NA,  4, NA, NA,  4,  3,
  "Robyn",         1, NA, NA, NA,  4,  4
)

# To pull live from the Assignment 2A PostgreSQL database instead
# (same connection pattern as week02_2a_approach.qmd):
# library(DBI); library(RPostgres)
# con <- dbConnect(RPostgres::Postgres(),
#   host = Sys.getenv("PGHOST", "localhost"), port = as.integer(Sys.getenv("PGPORT", "5432")),
#   dbname = Sys.getenv("PGDATABASE", "movie_ratings"), user = Sys.getenv("PGUSER", "postgres"),
#   password = Sys.getenv("PGPASSWORD"))
# ratings_long <- dbGetQuery(con, "SELECT u.name AS user, m.title AS movie, r.rating
#                             FROM ratings r JOIN users u ON r.user_id = u.user_id
#                             JOIN movies m ON r.movie_id = m.movie_id") |> filter(!is.na(rating))
# dbDisconnect(con)

ratings_wide
```

### Reshape to Long Format

```{r}
#| label: reshape-long

ratings_long <- ratings_wide |>
  pivot_longer(-user, names_to = "movie", values_to = "rating") |>
  filter(!is.na(rating))

ratings_long
```

### Global Mean

```{r}
#| label: global-mean

mu <- mean(ratings_long$rating)
mu
```

### Item Bias

```{r}
#| label: item-bias

item_bias <- ratings_long |>
  group_by(movie) |>
  summarize(b_i = mean(rating) - mu, n_ratings = n(), .groups = "drop")

item_bias
```

### User Bias

```{r}
#| label: user-bias

user_bias <- ratings_long |>
  left_join(item_bias, by = "movie") |>
  group_by(user) |>
  summarize(b_u = mean(rating - mu - b_i), n_ratings = n(), .groups = "drop")

user_bias
```

### Baseline Predictions

Predictions for every (user, movie) pair, clipped to the 1–5 scale, joined back against the actual rating where one exists.

```{r}
#| label: baseline-predictions

baseline_predictions <- expand_grid(
  user  = unique(ratings_long$user),
  movie = unique(ratings_long$movie)
) |>
  left_join(user_bias |> select(user, b_u), by = "user") |>
  left_join(item_bias |> select(movie, b_i), by = "movie") |>
  mutate(
    predicted_rating = pmin(pmax(mu + b_u + b_i, 1), 5)
  ) |>
  left_join(ratings_long, by = c("user", "movie")) |>
  rename(actual_rating = rating)

baseline_predictions
```

## Validation

### Fit Against Known Ratings

Since baseline estimate has no held-out test set built into a 7-person survey, the standard check is in-sample fit: how close the baseline prediction lands to each rating that's actually known, measured as RMSE.

```{r}
#| label: validate-rmse

known <- baseline_predictions |> filter(!is.na(actual_rating))

rmse <- sqrt(mean((known$predicted_rating - known$actual_rating)^2))
rmse
```

### Manual Spot Check

Hand-verify one prediction outside the pipeline: Eduardo's predicted rating for *Mean Girls* (a movie he didn't rate).

```{r}
#| label: validate-manual-check

manual_pred <- pmin(pmax(
  mu +
    user_bias$b_u[user_bias$user == "Eduardo"] +
    item_bias$b_i[item_bias$movie == "Mean Girls"],
  1), 5)

computed_pred <- baseline_predictions |>
  filter(user == "Eduardo", movie == "Mean Girls") |>
  pull(predicted_rating)

tibble(manual_pred, computed_pred, match = isTRUE(all.equal(manual_pred, computed_pred)))
```

## Recommendations

For each respondent, the top-ranked movies they haven't rated, by predicted baseline score.

```{r}
#| label: recommendations

recommendations <- baseline_predictions |>
  filter(is.na(actual_rating)) |>
  group_by(user) |>
  arrange(desc(predicted_rating), .by_group = TRUE) |>
  slice_head(n = 3) |>
  select(user, movie, predicted_rating) |>
  ungroup()

recommendations
```

## Visualization

```{r}
#| label: fig-baseline-heatmap
#| fig-width: 8
#| fig-height: 5

baseline_predictions |>
  mutate(known = !is.na(actual_rating)) |>
  ggplot(aes(x = movie, y = user, fill = predicted_rating)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = ifelse(known, as.character(actual_rating), round(predicted_rating, 1))),
            color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#4C72B0", high = "#C44E52", name = "Predicted\nrating") +
  labs(
    title = "Global Baseline Estimate: Predicted vs. Actual Ratings",
    subtitle = "Bold numbers = actual rating given · faint-background numbers = predicted (recommendation candidates)",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Discussion

With only 7 respondents and 6 movies (22 of 42 possible ratings actually collected), both `b_u` and `b_i` are estimated from very little data — a single rating from a respondent, or a single rating for a movie, can swing that bias substantially. The RMSE above should be read as a sanity check that the model fits the data it was trained on reasonably well, not as evidence of predictive accuracy on new ratings; a proper held-out evaluation would need many more responses than this survey collected.

## Team Note

This assignment was completed individually.

## Next Steps

The Video Explainer (due Monday) will walk through this code and the resulting recommendations.
