---
title: "Assignment 3A: Global Baseline Estimate — Approach"
subtitle: "DATA 607 · Week 3"
author: "Dillon Leeper"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
    css: styles.css

execute:
  warning: false
  message: false

editor: visual
---

## Overview

Most recommender systems rely on personalized algorithms such as content-based filtering or item-item collaborative filtering. The **Global Baseline Estimate** is a useful, simpler alternative: a non-personalized recommender that predicts a rating using only the overall rating distribution, how a given item tends to be rated relative to everyone else, and how a given user tends to rate relative to everyone else. It serves as a benchmark that any personalized algorithm should be expected to beat.

This document lays out the approach for implementing a Global Baseline Estimate recommender in R using the movie ratings survey data, following the algorithm outlined in the provided spreadsheet.

## The Algorithm

For a user $u$ and item (movie) $i$, the baseline prediction is:

$$
\text{baseline}(u, i) = \mu + b_u + b_i
$$

where:

-   $\mu$ — the global mean rating across all users and all movies
-   $b_i$ — the item bias: how much movie $i$'s average rating deviates from $\mu$
-   $b_u$ — the user bias: how much user $u$'s ratings deviate from $\mu$, after accounting for the items they rated

The result is typically clipped to the valid rating scale, since $\mu + b_u + b_i$ can fall outside the original bounds.

## Planned Approach

### 1. Data Preparation

Import the survey data and reshape it from wide format (one row per respondent, one column per movie) into long/tidy format with `tidyr::pivot_longer()`, so each row represents a single (user, movie, rating) observation. Missing/unrated cells will be filtered out rather than treated as zero ratings.

### 2. Compute the Global Mean

Calculate $\mu$ as the mean of every recorded rating in the long-format data frame, ignoring missing values.

### 3. Compute Item (Movie) Bias

For each movie, compute:

$$
b_i = \overline{\text{rating}_i} - \mu
$$

grouping by movie and averaging its ratings, then subtracting the global mean.

### 4. Compute User Bias

For each user, correct for item bias before averaging: subtract both $\mu$ and the relevant movie's $b_i$ from each of that user's ratings, then average the residuals:

$$
b_u = \text{mean}(\text{rating}_{u,i} - \mu - b_i)
$$

### 5. Compute Baseline Predictions

Combine $\mu$, $b_u$, and $b_i$ for every (user, movie) pair — including pairs the user has not yet rated — to produce a predicted rating, clipped to the valid scale (e.g., 1–5).

### 6. Generate Recommendations

For each user, rank the movies they have not yet rated by predicted baseline score and surface the top-ranked titles as recommendations.

## R Implementation Plan

The implementation will rely entirely on base R and `dplyr` — no modeling packages are needed, since the point of a baseline estimate is to be a simple, interpretable benchmark.

```{r}
#| label: setup
#| eval: false

library(tidyverse)
library(readxl)

ratings_wide <- read_excel("MovieRatings.xlsx")
```

```{r}
#| label: reshape-long
#| eval: false

ratings_long <- ratings_wide |>
  pivot_longer(
    cols = -user,
    names_to = "movie",
    values_to = "rating"
  ) |>
  filter(!is.na(rating))
```

```{r}
#| label: global-mean
#| eval: false

mu <- mean(ratings_long$rating, na.rm = TRUE)
```

```{r}
#| label: item-bias
#| eval: false

item_bias <- ratings_long |>
  group_by(movie) |>
  summarize(b_i = mean(rating, na.rm = TRUE) - mu)
```

```{r}
#| label: user-bias
#| eval: false

user_bias <- ratings_long |>
  left_join(item_bias, by = "movie") |>
  group_by(user) |>
  summarize(b_u = mean(rating - mu - b_i, na.rm = TRUE))
```

```{r}
#| label: baseline-predictions
#| eval: false

baseline_predictions <- expand_grid(
  user = unique(ratings_long$user),
  movie = unique(ratings_long$movie)
) |>
  left_join(user_bias, by = "user") |>
  left_join(item_bias, by = "movie") |>
  mutate(
    predicted_rating = mu + b_u + b_i,
    predicted_rating = pmin(pmax(predicted_rating, 1), 5)
  )
```

## Next Steps

The Code Base deliverable (due Sunday) will implement the steps above against the actual survey data, validate predictions against a held-out set of known ratings, and produce a ranked recommendation list per respondent. The Video Explainer (due Monday) will walk through the resulting code and findings.
