---
title: "Assignment 3B: Window Functions — Approach"
subtitle: "DATA 607 · Week 3"
author: "Dillon Leeper"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
    css: styles.css

execute:
  warning: false
  message: false

editor: visual
---

## Overview

This assignment calls for a dataset containing time series for two or more separate items, and the use of window functions (in SQL or `dplyr`) to calculate a year-to-date (YTD) average and a six-day moving average for each item. This document outlines the planned dataset and the window-function approach in both `dplyr` and SQL.

## Dataset

Daily closing prices for two to three tickers since January 1, 2022, pulled via `tidyquant::tq_get()` or `quantmod::getSymbols()`, both of which source free daily data from Yahoo Finance and support multiple symbols and a date range in a single call. This naturally satisfies the "two or more separate items" requirement, with each ticker forming its own time series. A synthetic, LLM-generated dataset is a fallback if API access is unreliable.

The working shape after retrieval will be a long data frame with columns `date`, `ticker`, and `close`.

## Planned Approach

### Year-to-Date Average

Group by ticker and calendar year, order by date within each group, and take a running (cumulative) mean that resets at the start of each January. This is a running window that grows one row at a time from the first trading day of the year through the current row.

### Six-Day Moving Average

Group by ticker only (not by year, since the window should roll across year boundaries), order by date, and take a trailing average over the current row plus the five prior rows — a fixed-width window that slides forward one row at a time.

## R Implementation Plan

```{r}
#| label: setup
#| eval: false

library(tidyverse)
library(tidyquant)
library(slider)

tickers <- c("AAPL", "MSFT", "BTC-USD")

prices <- tq_get(
  tickers,
  from = "2022-01-01",
  get  = "stock.prices"
) |>
  transmute(ticker = symbol, date, close)
```

```{r}
#| label: ytd-average
#| eval: false

prices_ytd <- prices |>
  group_by(ticker, year = year(date)) |>
  arrange(date, .by_group = TRUE) |>
  mutate(ytd_avg = cummean(close)) |>
  ungroup()
```

```{r}
#| label: six-day-moving-average
#| eval: false

prices_ma <- prices_ytd |>
  group_by(ticker) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    ma6 = slide_dbl(close, mean, .before = 5, .complete = TRUE)
  ) |>
  ungroup()
```

An equivalent approach using `zoo::rollmean()` in place of `slider::slide_dbl()`:

```{r}
#| label: six-day-moving-average-zoo-alt
#| eval: false

library(zoo)

prices_ma_alt <- prices_ytd |>
  group_by(ticker) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    ma6 = rollmean(close, k = 6, align = "right", fill = NA)
  ) |>
  ungroup()
```

## SQL Equivalent

If implemented in SQL instead of (or alongside) `dplyr`, the same two calculations map to standard window function syntax:

```sql
SELECT
  ticker,
  date,
  close,
  AVG(close) OVER (
    PARTITION BY ticker, EXTRACT(YEAR FROM date)
    ORDER BY date
    ROWS UNBOUNDED PRECEDING
  ) AS ytd_avg,
  AVG(close) OVER (
    PARTITION BY ticker
    ORDER BY date
    ROWS BETWEEN 5 PRECEDING AND CURRENT ROW
  ) AS ma_6day
FROM prices
ORDER BY ticker, date;
```

## Team Note

This assignment allows working in a small team; this submission will note whether the work was completed solo or with collaborators once that's finalized.

## Next Steps

The Code Base deliverable (due Sunday) will execute the plan above against the retrieved price data, validate the YTD and moving-average calculations against a few hand-checked rows, and visualize both series per ticker. The Video Explainer (due Monday) will walk through the resulting code and output.
